State-space analysis of a continuous gravitational wave source with a pulsar timing array: inclusion of the pulsar terms

### 2024-01-08 manuscript/manuscript_2/paper.tex
all minor comments addressed

Figure \ref{fig:corner_plot_1} displays results for the seven parameters in $\boldsymbol{\theta}_{\rm gw}$ for ten arbitrary noise realisations in the form of a traditional corner plot. The histograms are the one-dimensional posteriors for each parameter, marginalized over the six other parameters. Each coloured curve is the result for different noise realisation. The solid orange line marks the known injected value. The two-dimensional contours mark the (0.5, 1, 1.5, 2)-sigma level surfaces. All histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values. There is scant evidence of railing against the prior bounds. There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours, although weak-correlations are evident between $\Omega$ and $\Phi_0$, $\psi$ and $\alpha$, and $\iota$ and $h_0$. The injected value is contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 60 out of the 70 possible combinations of the seven parameters and 10 realizations (i.e. 86$\%$). The injected value is contained within the 90\% credible interval for all posterior realisations of $\Omega, \Phi_0, \psi$, and  $\iota$. The injected value is contained within the 5\% and 95\% quantiles for all of the one-dimensional marginalized posteriors. There is appreciable natural random dispersion in the one-dimensional posterior medians between noise realisations. The minimum coefficient of variation is  0.2\% for $\Omega$ , the maximum value is 52\% for $\Phi_0$. The remaining parameters have typical coefficients $\lessapprox 10 \%$. For this representative example, with characteristic strain $h_0 = 5 \times 10^{-15}$ the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately. Analogous results for synthetic data with higher strain ($h_0 = 10^{-12}$) are presented in Appendix \ref{sec:app_high_SNR}, Figure \ref{fig:corner_high_snr_appendix}. At higher strains the dispersion between noise realisations is reduced and the one dimensional posteriors for $\iota$ and $h_0$ are more symmetric about the injection value. \newline

In addition to $\boldsymbol{\theta}_{\rm gw}$, it is of interest how well the pulsar-term phase correction parameters $\chi^{(n)}$ can be estimated, since this parameter is a key feature of our approach. Figure \ref{fig:corner_plot_3} displays the results for a representative subset of five parameters of $\chi^{(n)}$, $ \{\chi^{(1)} \dots \chi^{(5)} \}$ over the ten noise realisations, analogously to Figure \ref{fig:corner_plot_1}. We do not display the resulting corner plot for all elements of $\chi^{(n)}$ because it is too big. With the exception of the results for $\chi^{(2)}$ (discussed below), all histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values. There is no evidence for correlations between parameter pairs. The majority of the elements of $\chi^{(n)}$ are recovered unambiguously across all noise realisations; for the 47 elements of $\chi^{n}$ (i.e. not just the five elements plotted in Figure \ref{fig:corner_plot_3}), the injected value is  contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 383 out of the 470 possible combinations of the 47 parameters and ten realizations (i.e. $\sim$ 81 $\%$). There are some elements of $\chi^{(n)}$ for which consistent estimates are not obtained across multiple realisations, e.g $\chi^{(2)}$ (third row, third column of Figure \ref{fig:corner_plot_3}). This corresponds to the phase-correction term for PSR J0340+4130. Out of all the pulsars in the array, this pulsar lies in the closest direction to our synthetic GW source, such that $\boldsymbol{n} \cdot \boldsymbol{q}^{(2)} = -0.96$. As a result the magnitude of $\cos \left( \chi^{(2)}\right) \sim 0.07$, and the influence of the GW is minimal ($g^{(2)}(t) \to 1$) The inability to accurately infer $\chi^{(2)}$ is contingent on the magnitude of the GW signal relative to the measurement noise; for synthetic data with a larger strain $h_0 = 10^{-12}$, $\chi^{(2)}$ can be recovered accurately; see Figure \ref{fig:corner_high_snr_appendix_chi} in Appendix \ref{sec:app_high_SNR}. \newline

Similar results are obtained for the remaining 4$N$ parameters of $\boldsymbol{\theta}_{\rm psr}'$; the injected values are recovered unambiguously across the nine noise realisations. For the sake of brevity we do not display the calculated posterior distributions here because the inference of $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this paper and PTA analyses. \newline

@@

\subsection{Extension to broader parameter space} \label{sec:parameter_space}
\textcolor{red}{TK: Figures 3 and 4 may be updated. Some additional simulations are currently running to double check some things. I expect the current figures to be representative of any updated results.} \newline 
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/pp_plot_canonical500}
	\caption{Fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself (i.e. PP plot). The injections are 200 simulated GW sources generated by drawing randomly five parameters in $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions in Table \ref{tab:parameters_and_priors}. Each coloured curve corresponds to a different parameter (see legend). The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad respectively in order to maintain an approximately constant SNR. The grey shaded contours label the $1\sigma$, $2\sigma$ and 3$\sigma$ confidence intervals. For parameters with well estimated posteriors, the PP curve should fall along the diagonal of unit slope. $\Omega$ and $\alpha$ are generally well-estimated (i.e. the curves lie close to the unit diagonal). The remaining three parameters, $\psi$, $\phi_0$ and $\delta$ show modest evidence of being over-constrained and stray outside the shaded region.}
	\label{fig:parameter_space}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/pp_plot_canonical_nochi}
	\caption{Same as Figure \ref{fig:parameter_space} but now assuming that $\chi^{(n)}$ are known exactly \textit{a priori} (i.e.with a delta-function prior for the nested sampling inference). The excursions from the shaded region seen in Figure  \ref{fig:parameter_space} are reduced; nearly all parameters, with the exception of $\delta$, are contained within the shaded region. }
	\label{fig:parameter_space2}
\end{figure}

Section \ref{sec:pe_and_ms} focuses on a single representative system, summarised in Table \ref{tab:parameters_and_priors}. In this section we test the method for a range of systems, varying the SMBHB source parameters $\boldsymbol{\theta}_{\rm gw}$ through astrophysically relevant ranges. The aim of this section is to confirm that the analysis scheme can be applied across an astrophysically relevant parameter space and that the arbitrary choice of $\boldsymbol{\theta}_{\rm gw}$ in Section \ref{sec:pe_and_ms} is not particularly advantageous by accident. \newline

We analyse 200 injections constructed by fixing $h_0 = 5 \times 10^{-15}$, $\iota =1.0$ rad and drawing the remaining five elements of $\boldsymbol{\theta}_{\rm gw}$ randomly from the prior distributions defined in Table \ref{tab:parameters_and_priors}. We fix $h_0$ and $\iota$ in order to maintain an approximately constant signal-to-noise ratio (SNR) across the 200 injections. For each injection we compute the posterior distribution of ${\boldsymbol{\theta}}_{\rm gw}$. The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$ are as specified in Section \ref{sec:synthetic_data} and Table  \ref{tab:parameters_and_priors}. \newline

It is prohibitive to display corner plots analogous to Figure \ref{fig:corner_plot_1} for 200 injections and seven elements of $\boldsymbol{\theta}_{\rm gw}$. Therefore, in order to summarise the results we use a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}. A PP plot displays the fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself. In the ideal case of perfect recovery, the PP plot should be a diagonal line of unit slope. \newline

The results are displayed in Figure \ref{fig:parameter_space}. The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels for 200 injections. The curves for the five parameters generally exhibit the discussed linear behaviour. Some parameters are consistently well estimated across the parameter space; the blue curve for $\Omega$ is consistently contained within the 2$\sigma$ shaded region. Conversely some parameters, e.g. $\psi$, stray outside the $1\sigma$ bounded region and show evidence of being slightly over-constrained; there are more injections contained within lower-value credible intervals than would be expected statistically, and less injections contained within higher-value credible intervals. This occurs due to the low SNR of the injected GW signal and the varying sensitivity of the specific PTA configuration used to different regions of the sky. Manually inspecting the individual corner plots for each of the 200 injections, it is evident that for a minority of SMBHB sources which lie in an unfavourable direction ($\boldsymbol{n} \cdot \boldsymbol{q}^{(n)} = -1$) the method either does not converge to a consistent unimodal one-dimensional posterior, or else a posterior is converged to but the estimate is suboptimal. For these sources their location with respect to the PTA makes it difficult to infer some elements of $\chi^{(n)}$ accurately, which impacts the inference of the other elements of $\boldsymbol{\theta}_{\rm gw}$. All injected sources are ``observed" with the same settings such as $T_{\rm obs}$, $T_{\rm cad}$, $n_{\rm live}$ etc. Some sources in unfavourable sky locations  may only be resolvable with different observation settings; we defer an in-depth exploration of sensitivity across the parameter space to a future work. In Figure \ref{fig:parameter_space2} we display a second PP plot, arranged identically to Figure \ref{fig:parameter_space}, but now we assume that the injected values of $\chi^{(n)}$ are known exactly, i.e. setting a delta-function prior on $\chi^{(n)}$ during the nested sampling inference. Now the excursions out of the shaded region are almost completely reduced. This indicates the importance of accurate estimates of $\chi^{(n)}$ and the sensitivity of a specific PTA configuration to different parts of the sky. In spite of the small excursions of some parameters in Figure \ref{fig:parameter_space}, it is reassuring that in this low-SNR regime all parameters generally follow the desired linear trend, and that key parameters such as $\Omega$ lie close to the unit diagonal.

\section{Earth terms vs pulsar terms}\label{sec:earth_vs_psr}
In Section \ref{sec:rep_example1} we have validated the state-space analysis scheme when the pulsar terms are included in the inference model. In this section, critically, we compare the effect of including the pulsar-term in the model (i.e. using Equation \eqref{eq:g_func_trig_chi}), as opposed to the omitting the pulsar-term (i.e. Equation \eqref{eq:g_func_trig_earth}). We do this using the specific representative case of the individual quasi-monochromatic SMBHB source described in Table \ref{tab:parameters_and_priors}. We refer to the model where the pulsar-term is included alongside the Earth term as $\mathcal{M}_{\rm psr \& earth}$. We refer to the model where the pulsar-term is omitted as 
$\mathcal{M}_{\rm earth}$. Model $\mathcal{M}_{\rm psr \& earth}$ is parameterised by $\boldsymbol{\theta}'$. Model $\mathcal{M}_{\rm earth}$ is parameterised by $\boldsymbol{\theta}_{\rm earth} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}'_{\rm psr}$. Both models are applied to identical realisations of the data, generated using the procedure as described in Section  \ref{sec:synthetic_data}. The priors on the static parameters in each model are also identical, with the addition of a uniform prior on $\chi^{(n)}$ for $\mathcal{M}_{\rm psr \& earth}$, as described in Appendix \ref{sec:set_priors}. \newline

>>>>

Figure \ref{fig:corner_plot_1} displays the posterior distribution of $\boldsymbol{\theta}_{\rm gw}$ for ten noise realisations in the form of a traditional corner plot for the representative SMBHB source parametrized as per the top portion of Table \ref{tab:parameters_and_priors}. The histograms are one-dimensional posteriors, marginalized over the six other parameters. Each coloured curve corresponds to a different noise realisation. The solid orange line marks the known injected value. The two-dimensional contours mark the (0.5, 1, 1.5, 2)-sigma level surfaces. All histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values. There is scant evidence of railing against the prior bounds. There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours, although weak-correlations are evident between $\Omega$ and $\Phi_0$, $\psi$ and $\alpha$, and $\iota$ and $h_0$. The injected value is contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 60 out of the 70 possible combinations of the seven parameters and 10 realizations. The injected value falls within the 90\% credible interval for $\Omega, \Phi_0, \psi$, and  $\iota$ 40 out of 40 times. There is appreciable natural dispersion in the one-dimensional posterior medians between realisations. The minimum and maximum coefficients of variation are 0.2\% and 52\% for $\Omega$ and $\Phi_0$ respectively. The remaining parameters have typical coefficients $\lessapprox 10 \%$. For this representative example, with characteristic strain $h_0 = 5 \times 10^{-15}$ the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately. Analogous results for synthetic data with higher strain ($h_0 = 10^{-12}$) are presented in Appendix \ref{sec:app_high_SNR} (see Figure \ref{fig:corner_high_snr_appendix}). The dispersion between noise realisations decreases with increasing $h_0$ is reduced and the one-dimensional posteriors for $\iota$ and $h_0$ are more symmetric about the injected value. \newline

\subsection{GW phase at each pulsar} \label{sec:chi_estim}

 It is important to check how well the pulsar-term phase $\chi^{(n)}$ is estimated, since the reparameterization involving $\chi^{(n)}$ is a key feature of our approach.

 approach, as explained in Section \ref{sec:pulsar_term2}. In short, it turns out that $\chi^{(n)}$ is estimated accurately for all $n$.

 Figure \ref{fig:corner_plot_3} displays the results as a corner plot for the representative subset $\chi^{(1)}, \dots, \chi^{(5)}$ over the ten noise realisations, in the same style as Figure \ref{fig:corner_plot_1}. We do not display the corner plot for $\chi^{(6)}$, $\dots$, $\chi^{(47)}$ because it is too big. With the exception of the results for $\chi^{(2)}$ (discussed below), all histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values. There is no evidence for correlations between parameter pairs. Most phases are recovered unambiguously across all noise realisations; for the 47 $\chi^{(n)}$ parameters (i.e. not just the five plotted in Figure \ref{fig:corner_plot_3}), the injected value is  contained within the 90\% credible interval of the one-dimensional marginalized posteriors in 383 out of the 470 possible combinations of the 47 parameters and ten realizations.

Rarely, $\chi^{(n)}$ is not estimated consistently across multiple realizations for some $n$, e.g $\chi^{(2)}$ (third row, third column of Figure \ref{fig:corner_plot_3}), corresponding to PSR J0340+4130. Out of all the pulsars in the array, this object matches most closely the direction to the synthetic GW source, with $\boldsymbol{n} \cdot \boldsymbol{q}^{(2)} = -0.96$,  $\cos \chi^{(2)} \approx 0.07$, and $g^{(2)}(t) \approx XYZ$ The inability to accurately infer $\chi^{(2)}$ improves with the signal-to-noise ratio (SNR); for $h_0 = 1 \times 10^{-12}$, $\chi^{(2)}$ is recovered accurately; see Figure \ref{fig:corner_high_snr_appendix_chi} in Appendix \ref{sec:app_high_SNR}. \newline

\subsection{Timing parameters} \label{sec:timing_parameters}
Similar results are obtained for the 4$N$ parameters in $\boldsymbol{\theta}_{\rm psr}'$ besides $\chi^{(n)}$. The injected values are recovered unambiguously across the 10 noise realisations. For the sake of brevity we do not display the calculated posterior distributions, because inferring $\boldsymbol{\theta}_{\rm gw}$ is the main focus of this paper and PTA analyses. In short, the estimates of $f_{\rm em}(t_1)$ and $\dot{f}_{\rm em}(t_1)$ are guided into narrow ranges by the narrow priors. [then add 1-2 sentences about sigma including giving some numbers on percentage accuracy (min and max)] We remind the reader that $\gamma^{(n)}$ is not estimated in this paper; typically it satisfies $\gamma^{(n)} \sim 10^{-5} T_{\rm obs}$ [refs from before], so its influence is muted. \newline

@@

\subsection{Exploring the SMBHB parameter domain} \label{sec:parameter_space} 
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/pp_plot_canonical500}
	\caption{Fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself (i.e. PP plot). The injections are 200 simulated GW sources generated by drawing randomly five parameters in $\boldsymbol{\theta}_{\rm gw}$ from the prior distributions in Table \ref{tab:parameters_and_priors}. Each coloured curve corresponds to a different parameter (see legend). The parameters $h_0$ and $\iota$ are fixed at $5 \times 10^{-15}$ and 1.0 rad respectively in order to maintain an approximately constant SNR. The grey shaded contours label the $1\sigma$, $2\sigma$ and 3$\sigma$ confidence intervals. For parameters with well estimated posteriors, the PP curve should fall along the diagonal of unit slope. $\Omega$ and $\alpha$ are generally well-estimated (i.e. the curves lie close to the unit diagonal). The remaining three parameters, $\psi$, $\phi_0$ and $\delta$ show modest evidence of being over-constrained and stray outside the shaded region.}
	\label{fig:parameter_space}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{images/pp_plot_canonical_nochi}
	\caption{Same as Figure \ref{fig:parameter_space} but now assuming that $\chi^{(n)}$ are known exactly \textit{a priori} (i.e.with a delta-function prior for the nested sampling inference). The excursions from the shaded region seen in Figure  \ref{fig:parameter_space} are reduced; nearly all parameters, with the exception of $\delta$, are contained within the shaded region. }
	\label{fig:parameter_space2}
\end{figure}

Section \ref{sec:rep_smbh_source} focuses on a single representative source, summarised in Table \ref{tab:parameters_and_priors}. In this section we test the method for a range of sources, varying $\boldsymbol{\theta}_{\rm gw}$. The aim is to verify that the analysis scheme works across an astrophysically relevant domain and that the arbitrary choice of $\boldsymbol{\theta}_{\rm gw}$ in Table \ref{tab:parameters_and_priors} is not advantageous by accident. \newline

We analyse 200 injections constructed by fixing $h_0 = 5 \times 10^{-15}$ and $\iota =1.0$ rad and drawing the remaining five elements of $\boldsymbol{\theta}_{\rm gw}$ randomly from the prior distributions defined in Table \ref{tab:parameters_and_priors}. We fix $h_0$ and $\iota$ in order to maintain an approximately constant SNR across the 200 injections. For each injection we compute the posterior distribution of ${\boldsymbol{\theta}}_{\rm gw}$. The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$ are as specified in Section \ref{sec:rep_example1} and Table  \ref{tab:parameters_and_priors}. \newline

It is prohibitive to display corner plots analogous to Figure \ref{fig:corner_plot_1} for 200 injections and seven elements of $\boldsymbol{\theta}_{\rm gw}$. Therefore, we summarise the results with the aid of a parameter-parameter (PP) plot \citep{doi:10.1198/106186006X136976}. A PP plot displays the fraction of injections included within a given credible interval of the estimated posterior, as a function of the credible interval itself. In the ideal case of perfect recovery, the PP plot should be a diagonal line of unit slope. \newline

The PP results are displayed in Figure \ref{fig:parameter_space}. The shaded grey contours enclose the $1\sigma$, $2\sigma$, and $3\sigma$ significance levels for 200 injections. The curves for the five parameters, colour-coded in the legend, are approximately linear. Some parameters are consistently well estimated across the parameter domain, e.g. the blue curve for $\Omega$ lies everywhere within the 2$\sigma$ shaded region. Conversely some parameters, e.g. $\psi$, stray outside the $1\sigma$ shaded region and show evidence of being slightly over-constrained; there are more injections contained within lower-value credible intervals than would be expected statistically, and fewer injections contained within higher-value credible intervals. This occurs due to the low SNR of the injected GW signal and the varying sensitivity of the specific PTA configuration used to different regions of the sky. \newline

Manually inspecting the individual corner plots for each of the 200 injections, we see that the nested sampler does not return a unimodal posterior in the rare event that the source is located unfavourably, with $\boldsymbol{n} \cdot \boldsymbol{q}^{(n)} \approx -1$, such that one cannot infer $\chi^{(n)}$ accurately (see Section \ref{sec:chi_estim}). All injected sources are ``observed" with the same settings such as $T_{\rm obs}$, $T_{\rm cad}$ and $n_{\rm live}$. In Figure \ref{fig:parameter_space2} we display a second PP plot, arranged identically to Figure \ref{fig:parameter_space}, but assuming that the injected values of $\chi^{(n)}$ are known exactly for the sake of testing, i.e. setting a delta-function prior on $\chi^{(n)}$. The excursions out of the shaded region diminish compared to Figure \ref{fig:parameter_space}, confirming the importance of accurate estimates of $\chi^{(n)}$.

%that for a minority of SMBHB sources which lie in an unfavourable direction ($\boldsymbol{n} \cdot \boldsymbol{q}^{(n)} = -1$) the method either does not converge to a consistent unimodal one-dimensional posterior, or else a posterior is converged to but the estimate is suboptimal.

%For these sources their location with respect to the PTA makes it difficult to infer some elements of $\chi^{(n)}$ accurately, which impacts the inference of the other elements of $\boldsymbol{\theta}_{\rm gw}$. 
%

%Some sources in unfavourable sky locations  may only be resolvable with different observation settings; we defer an in-depth exploration of sensitivity across the parameter space to a future work.

%
%and the sensitivity of a specific PTA configuration to different parts of the sky. In spite of the small excursions of some parameters in Figure \ref{fig:parameter_space}, it is reassuring that in this low-SNR regime all parameters generally follow the desired linear trend, and that key parameters such as $\Omega$ lie close to the unit diagonal.

\section{Bias when neglecting the pulsar terms}\label{sec:earth_vs_psr}
In Section \ref{sec:rep_example1} we have validated the state-space analysis scheme when the pulsar terms are included in the inference model. In this section, critically, we compare the effect of including the pulsar-term in the model (i.e. using Equation \eqref{eq:g_func_trig_chi}), as opposed to the omitting the pulsar-term (i.e. Equation \eqref{eq:g_func_trig_earth}). We do this using the specific representative case of the individual quasi-monochromatic SMBHB source described in Table \ref{tab:parameters_and_priors}. We refer to the model where the pulsar-term is included alongside the Earth term as $\mathcal{M}_{\rm psr \& earth}$. We refer to the model where the pulsar-term is omitted as 
$\mathcal{M}_{\rm earth}$. Model $\mathcal{M}_{\rm psr \& earth}$ is parameterised by $\boldsymbol{\theta}'$. Model $\mathcal{M}_{\rm earth}$ is parameterised by $\boldsymbol{\theta}_{\rm earth} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}'_{\rm psr}$. Both models are applied to identical realisations of the data, generated using the procedure as described in Section  \ref{sec:synthetic_data}. The priors on the static parameters in each model are also identical, with the addition of a uniform prior on $\chi^{(n)}$ for $\mathcal{M}_{\rm psr \& earth}$, as described in Appendix \ref{sec:set_priors}. \newline
